\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks

### 2015-10-15 nnt_manu.Rnw
edits to draft before friendly review

%% an abstract and keywords
\Abstract{
The supervised neural network has been applied as a machine learning technique to identify and predict emergent patterns in unstructured datasets.  A common criticism of the method is the inability to characterize relationships among variables from a fitted model.  Several methods proposed in recent years to `illuminate the black-box' have not been made available in an open-source programming environment.  This article describes the \pkg{NeuralNetTools} package that can be used for the interpretation of supervised neural network models created in \proglang{R}.  Functions in the package can be used to visualize a model using a neural network interpretation diagram, evaluate variable importance by disaggregating the model weights, and perform a sensitivity analysis of the response variables to changes in the input variables.  Methods are provided for objects from many of the common neural networks packages in \proglang{R}, including \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.  The article provides a brief overview of the theoretical foundation of neural networks, a description of the package structure and functions, and an applied example to provide a context for model development with \pkg{NeuralNetTools}.  Overall, the package provides a toolset for neural networks that complements existing quantitative techniques for data-intensive exploration.      
}
\Keywords{neural networks, plotnet, sensitivity, variable importance, \proglang{R}}
\Plainkeywords{neural networks, plotnet, sensitivity, variable importance, R} %% without formatting
%% at least one keyword must be supplied

@@

Statistical methods that have been used to support data exploration for inductive analysis are numerous \citep{Jain00}.  A common theme among these methods is the use of machine-learning algorithms where the primary objective is to identify emergent patterns in the data with minimal human intervention.  Neural networks, in particular, are designed to mimic the neuronal structure of the human brain by `learning' inherent data structures through adaptive algorithms \citep{Rumelhart86,Ripley96}.  Although the conceptual model was introduced several decades ago \citep{McCulloch43}, neural networks have had a central role in data intensive science.  The most popular form of neural network is the feed-forward \ac{mlp} trained using the backpropagation algorithm \citep{Rumelhart86}.  This model is typically used to predict the response of one or more variables given one to many explanatory variables.  The hallmark feature of the \ac{mlp} is the characterization of relationships using an arbitrary number of parameters (i.e., the hidden layer) that are chosen through an iterative training process with the backpropation algorithm.  Conceptually, the \ac{mlp} is nothing more than a hyper-parameterized non-linear model that can fit a smooth function to any dataset with almost non-existent residual error \citep{Hornik91}.

An arbitrarily large number of parameters to fit a neural network provides obvious predictive advantages, but conversely complicates the extraction of critical model information.  Information such as variable importance or model sensitivity are necessary aspects of exploratory data analysis that are not easily obtained from a neural network. As such, a common criticism is that neural networks are `black-boxes' that offer minimal insight into relationships among variables \citep[e.g.,][]{Paruelo97}.  \citet{Olden02} provide a rebuttal to this concern by describing methods to extract information from neural networks, most of which were previously available but not commonly used.  For example, \citet{Olden02} describe  the \ac{nid} for plotting \citep{Ozesmi99}, the Garson algorithm for variable importance \citep{Garson91}, and the Profile method for sensitivity analysis \citep{Lek96}.  These quantitative tools `illuminate the black box' by disaggregating the network parameters to characterize relationships between variables that are described by the model.  In essence, \ac{mlp} neural networks were developed for prediction but methods described in \citep{Olden02} leverage these models to describe data signals.  Increasing the accesibility of these diagnostic tools will have value for exploratory analysis in data science.

This article describes the \pkg{NeuralNetTools} package for \proglang{R} that was developed to improve the breadth and quality of information obtained from the \ac{mlp} neural network.  Functions provided by the package are those previously described in \citep{Olden02} but have not been available in an open-source programming environment.  The reach of the package is extensive such that generic functions were developed for model objects from the most popular neural network functions available in \proglang{R}.  The objecives of this article are to 1) provide an overview of the statistical foundation the \ac{mlp} network, 2) describe the theory and application of the main functions in the \pkg{NeuralNetTools} package, and 3) provide a context for appropriate use of neural networks and \pkg{NeuralNetTools} in data exploration using an applied example.  The package is currently available on the \ac{CRAN}, whereas the development version is maintained as a GitHub repository.

@@

An intriguing characteristic of neural networks is the relaxation of assumptions for the distributional characteristics of the response variables.  Unlike conventional approaches, neural networks have been popularized by their ability to model response variables with arbitrary distributions and can describe relationships in datasets with noisy or imprecise information.  The typical \ac{mlp} network is composed of multiple layers that define the transfer of information between input and response layers.  Information travels in one direction where a set of values for one to many variables in the input layer propagates through one or more hidden layers to the resulting layer of the response variables. `Hidden' layers between the input and response layers are key components of a neural network that mediate the transfer of information.  Just as the input and response layers are composed of variables or `nodes', each hidden layer is composed of nodes with weighted connections that define the strength of information flow between layers.  `Bias' layers connected to hidden and response layers may also be used that are analagous to intercept terms in a standard regression model.

Training a neural network model requires identifying the `optimal' weights that define the connections between the model layers.  The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.  Training is commonly achieved using the backpropagation algorithm described in detail in \citep{Rumelhart86}.  This algorithm identifies the optimal weighting scheme between layers through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.  The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.  The weights are then changed through a back-propagation step that begins by changing weights in the output layer and then the remaining hidden layers.  The process is repeated until the chosen error function is minimized.  Formulaically, a fitted \ac{mlp} neural network can be represented as \citep{Ripley96}:
\begin{equation}
y_k = f_k \left(\sum\limits_{j=1}^k w_{jk}f_j \left( \sum\limits_{i=1}^j w_{ij}x_i\right) \right)
\end{equation}
where the estimated value of the response variable $y_k$ is a summation of the product between multiple input variables $x$ and multiple hidden nodes $j$, as mediated by the respective weights $w$ and activation functions $f_j$ and $f_k$ for each hidden and output node as the information progresses through the network.

@@

Methods in \pkg{NeuralNetTools} were written for several \proglang{R} packages that can be used to create \ac{mlp} neural networks: \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}. Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.  Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the ``TAO-robust back-propagation algorithm'' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.  At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included \Sexpr{round(tots_nnt)}\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over \Sexpr{round(tots_nnet)}\%.  As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.  Methods are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class) objects.  Additional \code{default} or \code{numeric} methods are available for some of the generic functions.

@@

\pkg{NeuralNetTools} includes four main functions that were developed following techniques described in \citet{Olden02} and references therein.  These functions include \code{plotnet} to plot a neural network interpretation diagram, \code{garson} and \code{olden} functions to evaluate variable importance, and \code{lekprofile} for a sensitivity analysis of neural network response to input variables.  Most of the functions require the extraction of model weights in a common format for each of the neural network object classes in \proglang{R}.  The \code{neuralweights} function can be used to retrieve model weights for any of the model classes describe above.  A two-element \code{list} is returned with the first element describing the structure of the network and the second element as a named list of weight values for the input model.  The function is used internally within the main package functions but may be useful for comparing networks of different object classes.

A sample dataset is also included with \pkg{NeuralNetTools}.  The \code{neuraldat} dataset is a simple \code{data.frame} with 2000 rows of observations and columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}).  The input variables are random observations from a standard normal distribution and the response variables are linear combinations of the input variables with additional random components.  The response variables are also standardized from zero to one.  A common approach for data preprocessing prior to creating a neural network is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.  Novel datasets can be preprocessed to this common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.  The examples below use three models created from the \code{neuraldat} dataset and include \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{RSNNS}), and \code{nnet} (\pkg{nnet}) objects (note the syntax differences).

@@

The number of existing functions in \pkg{R} to view neural networks is minimal.  Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance. To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks in R.  Although useful for viewing the basic structure, the output is minimal and does not include extensive options for customization.

Tbe \code{plotnet} function in \pkg{NeuralNetTools} plots a network as a \acl{nid} \citep{Ozesmi99} and includes several options to customize plot aesthetics. A \ac{nid} is a modification to the standard conceptual illustration of the \ac{mlp} network that allows the user to view the thickness and color of the weight connections based on mangnitude and sign, respectively.  The default settings plot positive weights between layers as black lines and negative weights as grey lines. Line thickness is in proportion to absolute magnitude of each weight (\cref{fig:plotnet}).

<<plotnet, fig.show = 'hold', fig.height = 5.5, fig.width = 5.5, out.height = '7.5cm', out.width = '7.5cm', fig.cap = 'Examples from the \\code{plotnet} function showing neural networks as a standard graphic (\\ref{fig:plotnet1}) and using the \\acl{nid} (\\ref{fig:plotnet2}).  Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).  Default options are changed for the example.', fig.subcap = c('', '')>>=
par(mar = c(0, 0, 0, 0))
plotnet(mod3, nid = F, circle_col = 'grey', bord_col = 'grey')
plotnet(mod3)
@

A primary and skip layer network can also be plotted for \code{nnet} models with a skip layer connection. The default is to plot the primary network, whereas the skip layer network can be viewed with \code{skip = TRUE}. If \code{nid = TRUE}, the line widths for both the primary and skip layer plots are relative to all weights. Viewing both plots is recommended to see which network has larger relative weights. Plotting a network with only a skip layer (i.e., no hidden layer, \code{size = 0}) will include bias connections to the output layer, whereas these are not included in the plot of the skip layer if size is greater than zero.

<<plotnet_skip, fig.show = 'hold', fig.height = 5.5, fig.width = 5.5, out.height = '7.5cm', out.width = '7.5cm', fig.cap = 'Examples from the \\code{plotnet} function showing a neural network with a separate skip layer between the input and output layers.  The skip layer (\\ref{fig:plotnet_skip1}) and primary neural network (\\ref{fig:plotnet_skip2}) can be viewed separately with \\code{plotnet}.', results = 'hide', fig.subcap = c('', '')>>=
# create a model with a skip layer
modskip <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5, skip = TRUE)

@@

The \pkg{RSNNS} package provides several algorithms that can be used to prune connections or nodes in a neural network.  This approach can remove connection weights between layers or input nodes that do not contribute to the predictive performance of the network.  Pruning has the benefit of a potential increase in power by increasing the degrees of freedom or the identification of input nodes with low importance.  Algorithms in \pkg{RSNNS} for weight pruning include magnitude based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.  The \code{plotnet} function can be used to plot a pruned neural network, with options to omit or display the pruned connections (\cref{fig:plotprune}).

<<plotprune, eval = T, echo = T, fig.show = 'hold', fig.height = 7, fig.width = 7.7, out.width = '0.48\\textwidth', fig.cap = 'A pruned neural network using the ``Optimal Brain Surgeon\'\' algorithm described in \\citet{Zell98}.  The default plotting behavior of \\code{plotnet} is to omit pruned connections (\\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \\code{prune\\_col} argument (\\ref{fig:plotprune2}).', fig.subcap = c('', '')>>=
# pruned model using RSNNS
pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
  no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
  init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
 pruneFuncParams = pruneFuncParams)

@@

The primary benefit of visualizing a \ac{nid} with \code{plotnet} is the ability to evaluate network architecture and the variation in weighted connections between the layers.  Although useful as a general tool, the \ac{nid} is impractical for evaluating most networks given the amount of weighted connections.  Alternative methods to quantitatively describe a neural network deconstruct the model weights to determine variable importance, whereas similar information can only be qualitatively inferred from \code{plotnet}.  Two algorithms for evaluating variable importance are available in \pkg{NeuralNetTools}: Garson's algorithm for relative importance \citep{Garson91,Goh95} and Olden's conection weights algorithm \citep{Olden04}.

Garson's algorithm was originally described by \citet{Garson91} and further modified by \citet{Goh95}.  The \code{garson} function is an implementation of the method described in the appendix of \citet{Goh95}.  This method identifies the relative importance of each variable as an absolute magnitude by deconstructing all weighted connections between the layers in the network. For each input node, all weights connecting an input through the hidden layer to the response variable are identified to return a list of all weights specific to each input variable. A summed product of the connections for each input node are then scaled relative to all other inputs. A value for each input node indicates relative importance as the absolute magnitude from zero to one. The method is limited in that the direction of the response cannot be determined and only neural networks with one hidden layer and one output node can be evaluated.

The \code{olden} function is a more flexible approach to evaluate variable importance using the connection weights algorithm described in \citet{Olden04}. This method calculates importance as the product of the raw input-hidden and hidden-output connection weights between each input and output node and sums the product across all hidden nodes. An advantage is the relative contributions of each connection weight are maintained in both magnitude and sign, as compared to Garson's algorithm which only considers absolute magnitude. For example, connection weights that change sign (e.g., positive to negative) between the input-hidden to hidden-output layers would have a cancelling effect whereas Garson's algorithm may provide misleading results based on the absolute magnitude. An additional advantage is that Olden's algorithm is capable of evaluating neural networks with multiple hidden layers and response variables. The importance values assigned to each variable are also in units based on the summed product of the connection weights, whereas \code{garson} returns importance scaled from 0--1.

Both functions have similar implementations and require only a model object as input.  The default output is a \pkg{ggplot2} bar plot \citep[i.e., \code{geom\_bar},][]{Wickham09} that shows the relative importance of each input variable in the model (\cref{fig:plotimp}).  The plot aesthetics are based on internal code from the function but can be changed using conventional syntax for \pkg{ggplot2}.  The importance values can also be returned as a \code{data.frame} if \code{bar\_plot = FALSE}.

@@

\subfloat[][Model 3 (\code{nnet}), \code{garson}]{
<<plotimp-5, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
garson(mod3)
@
\label{fig:plotimp5}
}
\subfloat[][Model 3 (\code{nnet}), \code{olden}]{
<<plotimp-6, fig.height = 2, fig.width =4, out.width = '0.47\\textwidth', echo = F>>=
olden(mod3)
@
\label{fig:plotimp6}
}
\caption{Variable importance for three models using Garson's algorithm for relative importance (\code{garson}, \citealt{Garson91,Goh95}) and Olden's conection weights algorithm (\code{olden}, \citealt{Olden04}).  Garson's algorithm shows importance as absolute values from 0--1 whereas Olden's algorithm preserves sign and magnitude.  Importance values for Olden's algorithm are directly from the summed product of model weights and should only be evaluated based on the sign and relative magnitudes between variables.}
\label{fig:plotimp}
\end{figure}

@@

An alternative approach to evaluate variable interactions in a neural network is the Lek profile method, described briefly in \citet{Lek96} and in more detail in \citet{Gevrey03}. The profile method is fairly generic and can be extended to any statistical model in \pkg{R} with a \code{predict} method. However, it is one of few methods used to evaluate sensitivity in neural networks.  The profile method differs fundamentally from the variable importance algorithms by evaluating the behavior of response variables across different values of the input variables.

>>>>

%% an abstract and keywords
\Abstract{
Supervised neural networks have been applied as a machine learning technique to identify and predict emergent patterns in unstructured datasets.  A common criticism of the method is the inability to characterize relationships among variables from a fitted model. Although several methods proposed in recent years to `illuminate the black-box', they have not been made available in an open-source programming environment.  This article describes the \pkg{NeuralNetTools} package that can be used for the interpretation of supervised neural network models created in \proglang{R}.  Functions in the package can be used to visualize a model using a neural network interpretation diagram, evaluate variable importance by disaggregating the model weights, and perform a sensitivity analysis of the response variables to changes in the input variables.  Methods are provided for objects from many of the common neural networks packages in \proglang{R}, including \pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.  The article provides a brief overview of the theoretical foundation of neural networks, a description of the package structure and functions, and an applied example to provide a context for model development with \pkg{NeuralNetTools}.  Overall, the package provides a toolset for neural networks that complements existing quantitative techniques for data-intensive exploration.      
}
\Keywords{neural networks, plotnet, sensitivity, variable importance, \proglang{R}}
\Plainkeywords{neural networks, plotnet, sensitivity, variable importance, R} %% without formatting
%% at least one keyword must be supplied

@@

Statistical methods that have been used to support data exploration are numerous \citep{Jain00}.  A common theme among data intensive methods is the use of machine-learning algorithms where the primary objective is to identify emergent patterns with minimal human intervention.  Neural networks, in particular, are designed to mimic the neuronal structure of the human brain by `learning' inherent data structures through adaptive algorithms \citep{Rumelhart86,Ripley96}.  Although the conceptual model was introduced several decades ago \citep{McCulloch43}, neural networks have had a central role in emerging fields focused on data exploration.  The most popular form of neural network is the feed-forward \ac{mlp} trained using the backpropagation algorithm \citep{Rumelhart86}.  This model is typically used to predict the response of one or more variables given one to many explanatory variables.  The hallmark feature of the \ac{mlp} is the characterization of relationships using an arbitrary number of parameters (i.e., the hidden layer) that are chosen through iterative training with the backpropation algorithm.  Conceptually, the \ac{mlp} is a hyper-parameterized non-linear model that can fit a smooth function to any dataset with minimal residual error \citep{Hornik91}.

An arbitrarily large number of parameters to fit a neural network provides obvious predictive advantages, but conversely complicates the extraction of model information.  Diagnostic information such as variable importance or model sensitivity are necessary aspects of exploratory data analysis that are not easily obtained from a neural network. As such, a common criticism is that neural networks are `black-boxes' that offer minimal insight into relationships among variables \citep[e.g.,][]{Paruelo97}.  \citet{Olden02} provide a rebuttal to this concern by describing methods to extract information from neural networks.  Many of these methods were previously described but not commonly used.  For example, \citet{Olden02} describe  the \ac{nid} for plotting \citep{Ozesmi99}, the Garson algorithm for variable importance \citep{Garson91}, and the Profile method for sensitivity analysis \citep{Lek96}.  These quantitative tools `illuminate the black box' by disaggregating the network parameters to characterize relationships between variables that are described by the model.  Although \ac{mlp} neural networks were developed for prediction, methods described in \citet{Olden02} leverage these models to describe data signals.  Increasing the accesibility of these diagnostic tools will have value for exploratory data analysis and may also inform causal inference.

This article describes the \pkg{NeuralNetTools} package for \proglang{R} that was developed to improve the breadth and quality of information obtained from the \ac{mlp} neural network.  Functions provided by the package are those described in \citet{Olden02} but have not been previously available in an open-source programming environment.  The reach of the package is extensive such that generic functions were developed for model objects from the most popular neural network packages available in \proglang{R}.  The objectives are to 1) provide an overview of the statistical foundation of the \ac{mlp} network, 2) describe the theory and application of the main functions in the \pkg{NeuralNetTools} package, and 3) provide an applied example using neural networks and \pkg{NeuralNetTools} in data exploration.  The package is currently available on the \ac{CRAN}, whereas the development version is maintained as a GitHub repository.

@@

The typical \ac{mlp} network is composed of multiple layers that define the transfer of information between input and response layers.  Information travels in one direction where a set of values for variables in the input layer propagates through one or more hidden layers to the final layer of the response variables. `Hidden' layers between the input and response layers are key components of a neural network that mediate the transfer of information.  Just as the input and response layers are composed of variables or `nodes', each hidden layer is composed of nodes with weighted connections that define the strength of information flow between layers.  `Bias' layers connected to hidden and response layers may also be used that are analagous to intercept terms in a standard regression model.

Training a neural network model requires identifying the `optimal' weights that define the connections between the model layers.  The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.  Training is commonly achieved using the backpropagation algorithm described in \citep{Rumelhart86}.  This algorithm identifies the optimal weighting scheme through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.  The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.  The weights are then changed through a back-propagation step that begins by changing weights in the output layer and then the remaining hidden layers.  The process is repeated until the chosen error function is minimized, as in standard model-fitting techniques for regression \citep{Cheng94}.  Formulaically, a fitted \ac{mlp} neural network can be represented as \citep{Ripley96}:
\begin{equation}
y_k = f_k \left(\sum\limits_{j=1}^k w_{jk}f_j \left( \sum\limits_{i=1}^j w_{ij}x_i\right) \right)
\end{equation}
where the estimated value of the response variable $y_k$ is a summation of the product between multiple input variables $x$ and multiple hidden nodes $j$, as mediated by the weights $w$ and activation functions $f_j$ and $f_k$ for each hidden and output node.

@@

Methods in \pkg{NeuralNetTools} were written for several \proglang{R} packages that can be used to create \ac{mlp} neural networks: \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}. Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.  Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the ``TAO-robust back-propagation algorithm'' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.  At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included \Sexpr{round(tots_nnt)}\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over \Sexpr{round(tots_nnet)}\%.  As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.  Methods for each function are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class.  Additional \code{default} or \code{numeric} methods are available for some of the generic functions.

@@

\pkg{NeuralNetTools} includes four main functions that were developed following similar techniques in \citet{Olden02} and references therein.  Functions include \code{plotnet} to plot a neural network interpretation diagram, \code{garson} and \code{olden} functions to evaluate variable importance, and \code{lekprofile} for a sensitivity analysis of neural network response to input variables.  Most of the functions require the extraction of model weights in a common format for the neural network object classes in \proglang{R}.  The \code{neuralweights} function can be used to retrieve model weights for any of the model classes describe above.  A two-element \code{list} is returned with the first element describing the structure of the network (number of nodes in the input, hidden, and output layers) and the second element as a named list of model weights.  The function is used internally within the main functions but may be useful for comparing networks of different classes.

A common approach for data preprocessing is to normalize the input variables and to standardize the response variables \citep{Lek00,Olden02}.  A sample dataset that follows this format is included with \pkg{NeuralNetTools}.  The \code{neuraldat} dataset is a simple \code{data.frame} with 2000 rows of observations and five columns for two response variables (\code{Y1} and \code{Y2}) and three input variables (\code{X1}, \code{X2}, and \code{X3}).  The input variables are random observations from a standard normal distribution and the response variables are linear combinations of the input variables with additional random components.  The response variables are also standardized from zero to one.  Additional datasets can be preprocessed to this common format using the \code{scale} function from \pkg{base} and the \code{rescale} function from \pkg{scales} for the input and response variables, respectively.  The examples below use three models created from the \code{neuraldat} dataset and include \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{RSNNS}), and \code{nnet} (\pkg{nnet}) objects.

@@

Existing plot functions in \proglang{R} to view neural networks is minimal.  Such tools have practical use for visualizing network architecture and connections between layers that mediate variable importance. To our knowledge, only the \pkg{neuralnet} and \pkg{FCNN4R} packages provide plotting methods for \ac{mlp} networks in \proglang{R}.  Although useful for viewing the basic structure, the output is minimal and does not include extensive options for customization.

Tbe \code{plotnet} function in \pkg{NeuralNetTools} plots a \acl{nid} \citep[\acs{nid},][]{Ozesmi99} and includes several options to customize aesthetics. A \ac{nid} is a modification to the standard conceptual illustration of the \ac{mlp} network that changes the thickness and color of the weight connections based on magnitude and sign, respectively.  Positive weights between layers are shown as black lines and negative weights as grey lines. Line thickness is in proportion to absolute magnitude of each weight (\cref{fig:plotnet}).

<<plotnet, fig.show = 'hold', fig.height = 5.5, fig.width = 5.5, out.height = '7.5cm', out.width = '7.5cm', fig.cap = 'Examples from the \\code{plotnet} function showing neural networks as a standard graphic (\\ref{fig:plotnet1}) and using the \\acl{nid} (\\ref{fig:plotnet2}).  Labels outside of the nodes represent variable names and labels within the nodes indicate the layer and node (I: input, H: hidden, O: output, B: bias).', fig.subcap = c('', '')>>=
par(mar = c(0, 0, 0, 0))
plotnet(mod3, nid = F, circle_col = 'grey', bord_col = 'grey')
plotnet(mod3)
@

A primary and skip layer network can also be plotted for \code{nnet} models with a skip layer connection \citep{Ripley96}. Models with skip layers include additional connections from the input to output layers that bypass the hidden layer.  The default behavior of \code{plotnet} is to plot the primary network, whereas the skip layer can be viewed separately with \code{skip = TRUE}. If \code{nid = TRUE}, the line widths for both the primary and skip layer plots are relative to all weights. Plotting a network with only a skip layer (i.e., no hidden layer, \code{size = 0} in \code{nnet}) will include bias connections to the output layer, whereas these are included only in the primary plot if size is greater than zero.

<<plotnet_skip, fig.show = 'hold', fig.height = 5.5, fig.width = 5.5, out.height = '7.5cm', out.width = '7.5cm', fig.cap = 'Examples from the \\code{plotnet} function showing a neural network with a separate skip layer between the input and output layers.  The skip layer (\\ref{fig:plotnet_skip1}) and primary neural network (\\ref{fig:plotnet_skip2}) can be viewed separately with \\code{plotnet} by using \\code{skip = TRUE} or \\code{skip = FALSE}.', results = 'hide', fig.subcap = c('', '')>>=
# create a model with a skip layer
modskip <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5, skip = TRUE)

@@

The \pkg{RSNNS} package provides algorithms to prune connections or nodes in a neural network \citep{Bergmeir12}.  This approach can remove connection weights between layers or input nodes that do not contribute to the predictive performance of the network.  Pruning has additional benefits of an increase in power by increasing the degrees of freedom or the identification of input nodes with low importance.  Algorithms in \pkg{RSNNS} for weight pruning include magnitude based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.  The \code{plotnet} function can plot a pruned neural network, with options to omit or display the pruned connections (\cref{fig:plotprune}).

<<plotprune, eval = T, echo = T, fig.show = 'hold', fig.height = 7, fig.width = 7.7, out.width = '0.48\\textwidth', fig.cap = 'A pruned neural network from \\pkg{RSNNS} \\citep{Bergmeir12} using the ``Optimal Brain Surgeon\'\' algorithm described in \\citet{Zell98}.  The default plotting behavior of \\code{plotnet} is to omit pruned connections (\\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \\code{prune\\_col} argument (\\ref{fig:plotprune2}).', fig.subcap = c('', '')>>=
# pruned model using RSNNS
pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
  no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
  init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
 pruneFuncParams = pruneFuncParams)

@@

The primary benefits of visualizing a \ac{nid} with \code{plotnet} is the ability to evaluate network architecture and the variation in connections between the layers.  Although useful as a general tool, the \ac{nid} can be difficult to interpret given the amount of weighted connections in most networks.  Alternative methods to quantitatively describe a neural network deconstruct the model weights to determine variable importance, whereas similar information can only be qualitatively inferred from \code{plotnet}.  Two algorithms for evaluating variable importance are available in \pkg{NeuralNetTools}: Garson's algorithm for relative importance \citep{Garson91,Goh95} and Olden's conection weights algorithm \citep{Olden04}.

Garson's algorithm was originally described by \citet{Garson91} and further modified by \citet{Goh95}.  The \code{garson} function is an implementation of the method described in the appendix of \citet{Goh95} that identifies the relative importance of each variable as an absolute magnitude. For each input node, all weights connecting an input through the hidden layer to the response variable are identified to return a list of all weights specific to each input variable. A summed product of the connections for each input node are then scaled relative to all other inputs. A value for each input node indicates relative importance as the absolute magnitude from zero to one. The method is limited in that the direction of the response cannot be determined and only neural networks with one hidden layer and one output node can be evaluated.

The \code{olden} function is a more flexible approach to evaluate variable importance using the connection weights algorithm \citep{Olden04}. This method calculates importance as the summed product of the raw input-hidden and hidden-output connection weights between each input and output node. An advantage is the relative contributions of each connection weight are maintained in both magnitude and sign. For example, connection weights that change sign (e.g., positive to negative) between the input-hidden to hidden-output layers would have a cancelling effect, whereas Garson's algorithm may provide different results based on the absolute magnitude. An additional advantage is that Olden's algorithm can evaluate neural networks with multiple hidden layers and response variables. The importance values assigned to each variable are also in units based on the summed product of the connection weights, whereas \code{garson} returns importance scaled from 0--1.

Both functions have similar implementations and require only a model object as input.  The default output is a \pkg{ggplot2} bar plot \citep[i.e., \code{geom\_bar},][]{Wickham09} that shows the relative importance of each input variable in the model (\cref{fig:plotimp}).  The plot aesthetics are based on internal code that can be changed using conventional syntax for \pkg{ggplot2} applied to the output object.  The importance values can also be returned as a \code{data.frame} if \code{bar\_plot = FALSE}.

@@

\subfloat[][Model 3 (\code{nnet}), \code{garson}]{
<<plotimp-5, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
garson(mod3)
@
\label{fig:plotimp5}
}
\subfloat[][Model 3 (\code{nnet}), \code{olden}]{
<<plotimp-6, fig.height = 2, fig.width =4, out.width = '0.47\\textwidth', echo = F>>=
olden(mod3)
@
\label{fig:plotimp6}
}
\caption{Variable importance for three models using Garson's algorithm for relative importance (\code{garson}, \cref{fig:plotimp1,fig:plotimp3,fig:plotimp5}, \citealt{Garson91,Goh95}) and Olden's conection weights algorithm (\code{olden}, \cref{fig:plotimp2,fig:plotimp4,fig:plotimp6}, \citealt{Olden04}).  Garson's algorithm shows importance as absolute values from 0--1, whereas Olden's algorithm preserves sign and magnitude.  Importance values for Olden's algorithm are from the summed product of model weights and are not rescaled.}
\label{fig:plotimp}
\end{figure}

@@

An alternative approach to evaluate variable interactions in a neural network is the Lek profile method, described briefly in \citet{Lek96} and in more detail in \citet{Gevrey03}. The profile method differs fundamentally from the variable importance algorithms by evaluating the behavior of response variables across different values of the input variables. The method is generic and can be extended to any statistical model in \pkg{R} with a \code{predict} method. However, it is one of few methods used to evaluate sensitivity in neural networks.

The \code{lekprofile} function evaluates the effects of input variables by returning a plot of model predictions across the range of values for each variable.  The remaining explanatory variables are held constant when evaluating the effects of each input variable.  The \code{lekprofile} function provides two options for setting constant values of unevaluated explanatory variables.  The first option follows the original profile method by holding unevaluated variables at different quantiles (e.g., minimum, 20th percentile, maximum, \cref{fig:plotlek_sens1,fig:plotlek_bars1}). This is implemented by creating a matrix where the number of rows is the number of observations in the original dataset and the number of columns is the number of explanatory variables. All explanatory variables are held at a chosen constant value (e.g., median), while the variable of interest is sequenced from its minimum to maximum value across the range of observations. This matrix is then used to predict values of the response variable from a fitted model object. This is repeated for each explanatory variable to obtain all response curves.  Constant values are set in \code{lekprofile} by passing one or more values in the range 0--1 to the \code{group\_vals} argument.  The default holds variables at the minimum, 20\textsuperscript{th}, 40\textsuperscript{th}, 60\textsuperscript{th}, 80\textsuperscript{th}, and maximum percentiles (i.e., \code{group\_vals = c(0, 0.2, 0.4, 0.6, 0.8, 1)}).
